---
title: Early Life Gut Microbiome in Children following Spontaneous Preterm Birth and
  Maternal Preeclampsia
author: "Iskander Shadid"
date: "4/11/2023"
output:
  html_document: default
  pdf_document: default
---
  
```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.ext = '.jpeg')
```

#Required packages

```{r libraries}
library(Maaslin2)
library(phyloseq)
library(vegan)

library(ComplexUpset)
library(magrittr)
library(dplyr)
library(plyr)
library(grid)
library(gridExtra)
library(ggplot2)
library(ggpubr)
library(tableone)
library(Rmisc)
library(reghelper)

```

#Loading the VDAART data (dataframe and phyloseq) and wrangling 

```{r data}

# SECTION 1. VDAART dataset DATAFRAME  ----------------------------------------------------------

#### Microbial data ####
# Load microbe data sets collapes at genus level from new Illumina 16s data from QIIME2 pipeline at all time points (n=5) and full dataset (g_all)
load("/phyloseq_M3_tax.rda") # 3-6 months
load("/phyloseq_Y1_tax.rda") # 1 year
load("/phyloseq_Y3_tax.rda") # 3 year
load("/phyloseq_Y4_tax_genus.rda") # 4 year
load("/phyloseq_Y5_tax.rda") # 5 year
load("/FinalV2_g_all.rda") # all years together

# Unique ids used for this study
all_unique <- dfall[!duplicated(dfall$vid),]
all_unique <- all_unique[!is.na(all_unique$vid),]

#### Meta data ####
# load metadata 
db <- read.csv("/jama2016tab2.csv")
vdaartbirth <- read.csv("/vdaartbirth.csv")
PE_metadata <- read.csv("/peanalysis.csv", stringsAsFactors=FALSE, header=TRUE) #n=816
enroll <- read.csv("/EnrollMaternal.csv")
load("/OG806vids.rda") # vID's used in total VDAART #806

# merge all ids with metadata
tomerge1 <- merge(asthma.df, 
                  db[,c("vid","delgestage","m6breastfed","trmt")],
                  stringsAsFactors=FALSE, header=TRUE, 
                  by="vid", all.x = T) 

tomerge2 <- merge(tomerge1, 
                  vdaartbirth[,c("vid","modedelivery","comlabor","pihghtn","pprom","prom","pretermdelivery","antibiotic", "bweight")],
                  stringsAsFactors=FALSE, header=TRUE, 
                  by="vid", all.x = T) 

tomerge3 <- merge(tomerge2, 
                  PE_metadata[,c("vid","pefull")],
                  stringsAsFactors=FALSE, header=TRUE, 
                  by="vid", all.x = T) 

# final dataset 
all806 <- tomerge3

# modify metadata columns 

#race
all806$craceeth[all806$craceeth=="African American (Hispanic and Non-Hispanic)"]<-"Black, African American"
all806$craceeth[all806$craceeth=="Caucasian (Hispanic)"]<-"Caucasian"
all806$craceeth[all806$craceeth=="Caucasian (Non-Hispanic)"]<-"Caucasian"

# maternal education
all806$grad[all806$meducation=="College graduate"]<-"College or higher"
all806$grad[all806$meducation=="Graduate school"]<-"College or higher"
all806$grad[all806$meducation=="Technical school"]<-"Not a college graduate"
all806$grad[all806$meducation=="Graduated from high school"]<-"Not a college graduate"
all806$grad[all806$meducation=="Junior college/some college"]<-"Not a college graduate"
all806$grad[all806$meducation=="Did not graduate from high school"]<-"Not a college graduate"

# maternal income 
all806$income[all806$mincome=="$100,000 - $149,999"]<-"Over $50,000"
all806$income[all806$mincome=="$50,000 - $74,999"]<-"Over $50,000"
all806$income[all806$mincome=="$75,000 - $99,999"]<-"Over $50,000"
all806$income[all806$mincome=="Over $150,000"]<-"Over $50,000"
all806$income[all806$mincome=="$30,000 - $49,999"]<-"Less than $50,000"
all806$income[all806$mincome=="Less than $30,000"]<-"Less than $50,000"
all806$income[all806$mincome=="Do not know"]<-"Do not know/prefer not to answer"
all806$income[all806$mincome=="Prefer not to answer"]<-"Do not know/prefer not to answer"

# preeclampsia
all806$preeclampsia[all806$pefull==1]<-"Yes"
all806$preeclampsia[all806$pefull==0]<-"No"

# csection
all806$Csection <- ifelse(all806$modedelivery=="C-section", "Yes", "No")
table(all806$Csection)

# marital status 
all806$married[!all806$mmaritalstatus=="Married"]<-"Not Married (Separated or Divorced)"
all806$married[all806$mmaritalstatus=="Married"]<-"Married"

# sPTB <<<< 4 erroneous ID's <-- change to sPTB==No
all806$sPTB=ifelse(all806$pretermdelivery=="Yes" & all806$comlabor=="Spontaneous", "Yes","No")
all806$mage_30[all806$mage>=30]<-">=30"
all806$mage_30[all806$mage<30]<-"<30"
table(all806$sPTB)

#fix ID's
all806$sPTB <- ifelse(all806$sPTB=="Yes" & all806$preeclampsia== "Yes", "No", all806$sPTB)
table(all806$sPTB)
#
all806$pe_and_preterm=ifelse(all806$pefull==1 & all806$pretermdelivery=="Yes", "Yes","No")
all806$pe_no_preterm=ifelse(all806$pefull==1 & all806$pretermdelivery=="No", "Yes","No")
all806$antibiotic[all806$antibiotic=="Do not know"]<-NA

all806$sptb_and_pprom <- ifelse(all806$sPTB=="Yes" & all806$pprom=="Yes", "Yes","No")
all806$sptb_no_pprom <- ifelse(all806$sPTB=="Yes" & all806$pprom=="No", "Yes","No")

all806$GA_U37wks <- ifelse(all806$delgestagewks < 37 & all806$delgestagewks >= 34, "Yes", "No")
all806$GA_U34wks <- ifelse(all806$delgestagewks < 34, "Yes", "No")

# Mark included(n=671)/excluded(n=134)
all806$microbiome <- 0
all806$microbiome[all806$vid%in%all_unique$vid] <- 1

# missing to na 
all806$m6breastfed <- ifelse(all806$m6breastfed=="", NA, all806$m6breastfed)
all806$antibiotic[all806$antibiotic==""] <- NA

# separate variable for missing data to use in Table 1
all806$bfmis <- ifelse(is.na(all806$m6breastfed), "Yes", "No")
all806$antimis <- ifelse(is.na(all806$antibiotic), "Yes", "No")

# add birthweight category
all806$bwcat <- ifelse(all806$bweight>= 2500, "Normal", 
                       ifelse(all806$bweight <2500, "Underweight", NA))

#### Merge microbial and metadata ####
# Merge with microbes
db36 <- M3_FF %>% merge(y=all806, by="vid")
db1 <- Y1_FF %>% merge(y=all806, by="vid")
db3 <- Y3_FF %>% merge(y=all806, by="vid")
db4 <- Y4_FF %>% merge(y=all806, by="vid")
db5 <- Y5_FF %>% merge(y=all806, by="vid")

# Add column with exact age at stool sample collection for the infants at 3-6 months:

# Merge in ages at stool sample collection for infants 
stooldate <- read.csv("/ttrimdietm6ffsamp.csv")
x<-c("vid","childm3ffdate")
stooldate<-subset(stooldate, select=x)

# Merge in delivery date
epi1 <- read.csv("/babyracen819.csv")
x<-c("vid","deliverydate")
epi1<-subset(epi1, select=x)

# Merge delivery date in with stool collection date
meta <- merge(epi1, stooldate, by="vid", all.x=TRUE)

### for 7 subjects, birthdate and stool collection date missing
### use Excel sheet from Leanna to add these dates
meta$childm3ffdate <- as.character(meta$childm3ffdate)


# Create an age variable (stool sample collection date minus delivery date)
meta$stooldate<-as.Date(paste(meta$childm3ffdate), format = "%m/%d/%Y")
meta$ddate<-as.Date(meta$deliverydate, format = "%m/%d/%Y")
meta$age<-(meta$stooldate-meta$ddate)/365.25 # In years
meta$age<-as.numeric(meta$age)

# add age & time column
db36 <- db36 %>% merge(y=meta[,c("vid","age")], by="vid")
db1$age <- 1
db3$age <- 3
db4$age <- 4
db5$age <- 5

# save datasets 
#getwd()
#save(db36, file="finalV2_g_m3.rda")
#save(db1, file="finalV2_g_y1.rda")
#save(db3, file="finalV2_g_y3.rda")
#save(db4, file="finalV2_g_y4.rda")
#save(db5, file="finalV2_g_y5.rda")

# create one merged data set
M3Y1 <- rbind(db36, db1)
M3Y1Y3 <- rbind(M3Y1, db3)
M3Y1Y3Y4 <- rbind(M3Y1Y3, db4)
dfall <- rbind(M3Y1Y3Y4, db5)

# Remove entirely empty microbe columns if any
nonnum <- names(dfall[,c(1:3, 405:463)])
numcols <- names(dfall[,c(4:404)])
cols_to_keep <- colSums(dfall[,numcols]) != 0 
dfall <- dfall[, c(nonnum, numcols[cols_to_keep])]
# Remove ID's with zero counts for all microbes if any
del <- rowSums(dfall[63:437]) == 0
dfall <- dfall[!del, ]

# save dataset
#save(dfall, file="/FinalV2_g_all.rda")

# SECTION 2. VDAART dataset PHYLOSEQ  ----------------------------------------------------------

#### Microbial data ####
#load phyloseq objects 
data <- readRDS("/data_g.rds") # genus collapsed phyloseq from QIIME2_Pipeline

# pull rownames and make rowname column
sam_data(data)$rname <- rownames(sam_data(data))

# keep as separate reference dataframe to merge later
phylo_rows <- data.frame( sam_data(data)[,c("vid", "rname", "COLLECTION")])

#### Meta data ####
meta_phylo <- all806

# add exact age data
meta_phylo <- meta_phylo %>% merge(y=meta[,c("vid","age")], by="vid")

#### Merge microbial and metadata ####
# merge phyloseq rownames reference dataframe with metadata
df <- phylo_rows %>% merge(y=meta_phylo, by="vid")
rownames(df) <- df$rname

# adjust sampling time point category 
df$Time <- ifelse(df$COLLECTION == "Fecal Flora Collection month 3", "3-6 Months",
                   ifelse(df$COLLECTION == "Year 1 FF Collection", "1 Year",
                          ifelse(df$COLLECTION == "Year 3 FF Collection", "3 Years",
                                 ifelse(df$COLLECTION == "C4-Year 4 FF Collection", "4 Years",
                                        ifelse(df$COLLECTION == "C5-Year 5 FF Collection", "5 Years", NA)))))
df$Time <- factor(df$Time, levels = c("3-6 Months", "1 Year", "3 Years", "4 Years", "5 Years"))

# Turn into new `sample_data` 
sam.new <- sample_data(df)

# Merge with original phyloseq object
df <- merge_phyloseq(data, sam.new)

# save phyloseq object now including new metadata
#saveRDS(df, "/phylo_finalV2.rds")

```

# Table 1. Characteristics and comparisons of subgroups included in the study.  

```{r tab1}

# Select only inlcuded study subjects 
mb <- subset(all806, microbiome==1)
mb$cats <- ifelse(mb$preeclampsia=="Yes", "Preeclampsia", 
                  ifelse(mb$sPTB=="Yes", "sPTB","Normal"))
table(mb$cats)

# variables
vars=c("mage","sitename","craceeth", "antibiotic" , "antimis", "income","married", "cgender", "masthma", "preeclampsia", "pe_and_preterm", "pe_no_preterm","trmt", "sPTB","sptb_and_pprom","sptb_no_pprom","delgestagewks","GA_U37wks" , "GA_U34wks","Csection","m6breastfed", "bfmis","pretermdelivery","pihghtn","pprom", "grad", "bweight", "bwcat")

# create and print table 
tableOne <- CreateTableOne(vars = vars, strata = "cats", data = mb)
tableOneExport=print(tableOne, quote = F, noSpaces = TRUE, showAllLevels = T, na.rm=T)

```

# Table 2. MAASLIN2 analysis of main phenotypes 

```{r tab2}

# Factor child's race
dfall$craceeth <- factor(dfall$craceeth, levels = c("Black, African American",
                                                    "Caucasian",
                                                    "Other"))

prevalence_percentage <- colSums(dfall[63:437] != 0) / nrow(dfall) * 100
quantile(prevalence_percentage) 
#        0%         25%         50%         75%        100% 
#0.06031363  0.18094089  1.38721351 10.01206273 95.11459590 
# set to 0.2%

# run maaslin2 model ZINB with random effects for all time points (dfall)
# main phenotypes

Maaslin2(dfall[63:437],
         dfall[,c("vid", "preeclampsia","sPTB","antibiotic","Csection", "cgender", "m6breastfed", "sitename", "age", "craceeth")],
         "/DATA",
         min_abundance = 0.001,
         min_prevalence = 0.002,
         normalization = "TMM",
         transform = "NONE",
         analysis_method = "ZINB",
         max_significance = 0.7,
         random_effects =c("vid", "sitename"),
         fixed_effects =  c("preeclampsia","sPTB","antibiotic","Csection", "cgender", "m6breastfed", "age", "craceeth"),
         reference = c("craceeth,Black, African American"),
         correction = "BH",
         standardize = TRUE,
         cores = 1,
         plot_heatmap = TRUE,
         plot_scatter = TRUE,
         heatmap_first_n = 50)

```

# Table 3. MAASLIN2 analysis of SUB phenotypes 

```{r tab3}

# run maaslin2 model ZINB with random effects for all time points (dfall)
# subphenotypes
Maaslin2(
  dfall[63:437],
  dfall[,c("vid", "pe_and_preterm","pe_no_preterm","sptb_and_pprom","sptb_no_pprom",
           "antibiotic","Csection", "cgender", "m6breastfed", "sitename", "age", "craceeth")],
  "/Users/iskandershadid/Microbiome/Merged/all_wRace/Vid_site_secondary",
  min_abundance = 0.001,
  min_prevalence = 0.002,
  normalization = "TMM",
  transform = "NONE",
  analysis_method = "ZINB",
  max_significance = 0.7,
  random_effects =c("vid", "sitename"),
  fixed_effects =  c("pe_and_preterm","pe_no_preterm","sptb_and_pprom",
                     "sptb_no_pprom","antibiotic","Csection", "cgender", "m6breastfed", "age", "craceeth"),
  reference = c("craceeth,Black, African American"),
  correction = "BH",
  standardize = TRUE,
  cores = 1,
  plot_heatmap = TRUE,
  plot_scatter = TRUE,
  heatmap_first_n = 50)

```

# Figure 2. Infant gut microbiome diversity trajectories from birth to age 5 years. 

```{r fig2}
#### ALPHA DIVERSITY ####

## trim data 
# Months 3-6
M3 <- db36
nonnum <- names(M3[,c(1:3, 405:463)])
numcols <- names(M3[,c(4:404)])
cols_to_keep <- colSums(M3[,numcols]) != 0 
M3 <- M3[, c(nonnum, numcols[cols_to_keep])]

# year 1
Y1 <- db1
nonnum <- names(Y1[,c(1:3, 405:463)])
numcols <- names(Y1[,c(4:404)])
cols_to_keep <- colSums(Y1[,numcols]) != 0 
Y1 <- Y1[, c(nonnum, numcols[cols_to_keep])]

# year 3
Y3 <- db3
nonnum <- names(Y3[,c(1:3, 405:463)])
numcols <- names(Y3[,c(4:404)])
cols_to_keep <- colSums(Y3[,numcols]) != 0 
Y3 <- Y3[, c(nonnum, numcols[cols_to_keep])]

# year 4 
Y4 <- db4
nonnum <- names(Y4[,c(1:3, 405:463)])
numcols <- names(Y4[,c(4:404)])
cols_to_keep <- colSums(Y4[,numcols]) != 0 
Y4 <- Y4[, c(nonnum, numcols[cols_to_keep])]

# year 5
Y5 <- db5
nonnum <- names(Y5[,c(1:3, 405:463)])
numcols <- names(Y5[,c(4:404)])
cols_to_keep <- colSums(Y5[,numcols]) != 0 
Y5 <- Y5[, c(nonnum, numcols[cols_to_keep])]

# alpha diversity 
M3$alpha_shan <- diversity(M3[63:272],MARGIN = 1,index = "shannon") # 3-6 months
Y1$alpha_shan <- diversity(Y1[63:333],MARGIN = 1,index = "shannon") # year 1
Y3$alpha_shan <- diversity(Y3[63:355],MARGIN = 1,index = "shannon") # year 3
Y4$alpha_shan <- diversity(Y4[63:338],MARGIN = 1,index = "shannon") # year 4
Y5$alpha_shan <- diversity(Y5[63:311],MARGIN = 1,index = "shannon") # year 5

# Merge data form all time points now including alpha diversity
dfall <- rbind(M3[,c(1:62, 273)], Y1[,c(1:62,334)], 
               Y3[,c(1:62,356)], Y4[,c(1:62,339)], Y5[,c(1:62, 312)])

# Time point category
dfall$Time <- ifelse(dfall$age < 1, "3-6 Months",
                     ifelse(dfall$age == 1, "1 Year",
                            ifelse(dfall$age == 3, "3 Years",
                                   ifelse(dfall$age == 4, "4 Years",
                                          ifelse(dfall$age == 5, "5 Years", NA)))))
dfall$Time <- factor(dfall$Time, levels = c("3-6 Months", "1 Year", "3 Years", "4 Years", "5 Years"))
table(dfall$Time)


## Alpha Plots

# PE
# Standard error calculation PE
dfall_plot_pe_shan <- summarySE(dfall[dfall$preeclampsia=="Yes",], measurevar = "alpha_shan", groupvars = "Time")
dfall_plot_pe_shan$div <- "shan"
colnames(dfall_plot_pe_shan)[3] <- "alpha"
dfall_plot_pe_shan$Preeclampsia <- "Yes"

# Standard error calculation NO PE
dfall_plot_no_pe_shan <- summarySE(dfall[dfall$preeclampsia=="No",], measurevar = "alpha_shan", groupvars = "Time")
dfall_plot_no_pe_shan$div <- "shan"
colnames(dfall_plot_no_pe_shan)[3] <- "alpha"
dfall_plot_no_pe_shan$Preeclampsia <- "No"

# Bind PE 
dfall_plot_pe <- rbind(dfall_plot_pe_shan, dfall_plot_no_pe_shan)
dfall_plot_pe <- dfall_plot_pe[!is.na(dfall_plot_pe$Time),]

# SPTB
# Standard error calculation SPTB
dfall_plot_sptb_shan <- summarySE(dfall[dfall$sPTB=="Yes",], measurevar = "alpha_shan", groupvars = "Time")
dfall_plot_sptb_shan$div <- "shan"
colnames(dfall_plot_sptb_shan)[3] <- "alpha"
dfall_plot_sptb_shan$sPTB <- "Yes"

# Standard error calculation NO SPTB
dfall_plot_no_sptb_shan <- summarySE(dfall[dfall$sPTB=="No",], measurevar = "alpha_shan", groupvars = "Time")
dfall_plot_no_sptb_shan$div <- "shan"
colnames(dfall_plot_no_sptb_shan)[3] <- "alpha"
dfall_plot_no_sptb_shan$sPTB <- "No"

# Bind SPTB
dfall_plot_sptb <- rbind(dfall_plot_sptb_shan, dfall_plot_no_sptb_shan)
dfall_plot_sptb <- dfall_plot_sptb[!is.na(dfall_plot_sptb$Time),]


# Plot alpha diversity PE
pe <- ggplot(dfall_plot_pe[dfall_plot_pe$div=="shan",], aes(x=Time, y=alpha, col=Preeclampsia, group=Preeclampsia))+
  geom_errorbar(aes(ymin=alpha-se, ymax=alpha+se), width =.1)+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(tag = "(A)")+
  xlab("\nSampling Time")+
  ylab("Shannon Index\n")+
  scale_color_manual(values=c("#386CB0","#F0027F"),
                     name="Preeclampsia",
                     breaks=c("No","Yes"),
                     labels=c("No","Yes"))+
  theme_classic()+
  theme(legend.position=c(0.8,0.3),
        plot.tag.position =  c(0.01,1.1),
        plot.tag = element_text(face="bold"),
        text=element_text(size=14))

# Plot alpha diversity SPTB
sptb <- ggplot(dfall_plot_sptb[dfall_plot_sptb$div=="shan",], aes(x=Time, y=alpha, col=sPTB, group=sPTB))+
  geom_errorbar(aes(ymin=alpha-se, ymax=alpha+se), width =.1)+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(tag = "(B)")+
  xlab("\nSampling Time")+
  ylab("Shannon Index\n")+
  scale_color_manual(values=c("#7FC97F","#FDC086"),
                     name="Spontaneous\nPreterm Birth",
                     breaks=c("No","Yes"),
                     labels=c("No","Yes"))+
  theme_classic()+
  theme(legend.position=c(0.8,0.35),
        text=element_text(size=14),
        plot.tag.position =  c(0.01,1.1),
        plot.tag = element_text(face="bold"),
        legend.background = element_rect(fill="transparent"))

# Arrange plots 
plot_alpha <- ggarrange(pe, sptb, nrow=2, ncol=1)

#### Beta Diversity ##### 

# Remove 0 IDs & Maternal samples
df <- subset_samples(df, sample_sums(df) > 0)
df <- subset_samples(df, COLLECTION!="Mother FF Third Trimester")

# Split data by time points
age0.5 <- subset_samples(df, Time == "3-6 Months")
age1 <- subset_samples(df, Time == "1 Year")
age3 <- subset_samples(df, Time == "3 Years")
age4 <- subset_samples(df, Time == "4 Years")
age5 <- subset_samples(df, Time == "5 Years")

# Split data by phenotypes
pe.df <-  subset_samples(df, preeclampsia=="Yes")
sptb.df <-  subset_samples(df, sPTB=="Yes")
health <-  subset_samples(df, preeclampsia=="No" & sPTB == "No")

## Create PCOA plot

# Ordinate all subjects
data_all<- ordinate(
  physeq = df, 
  method = "PCoA", 
  distance = "bray")

# switch PE axis (*-1)
data_all[c("vectors")] <- lapply(data_all[c("vectors")], 
                                 function(x) {x[,"Axis.1"] <- x[,"Axis.1"] * -1
                                 x})
# Ordinate PE subjects
data_pe<- ordinate(
  physeq = pe.df, 
  method = "PCoA", 
  distance = "bray")

# Ordinate SPTB subjects
data_sptb<- ordinate(
  physeq = sptb.df, 
  method = "PCoA", 
  distance = "bray")

# switch Pe axis (*-1)
data_sptb[c("vectors")] <- lapply(data_sptb[c("vectors")], 
                                  function(x) {x[,"Axis.1"] <- x[,"Axis.1"] * -1
                                  x})
# Ordinate control (healthy) subjects
data_health<- ordinate(
  physeq = health, 
  method = "PCoA", 
  distance = "bray")

# switch Pe axis (*-1)
data_health[c("vectors")] <- lapply(data_health[c("vectors")], 
                                    function(x) {x[,"Axis.1"] <- x[,"Axis.1"] * -1
                                    x})

# Plot PCoA with age color-coded ALL Subjects
plotall <- plot_ordination(
  physeq = df,
  ordination = data_all,
  color = "COLLECTION",
  title = "Fecal Microbiome Beta Diversity\n(Bray-Curtis Dissimilarity) Over Early Life Course")+ 
  scale_color_manual(values = c("#FDE725FF","#95D840FF","#1F968BFF","darkorchid4", "#FF5F15" ),
                     name="Age at Stool Sample\nCollection",
                     breaks=c("Fecal Flora Collection month 3","Year 1 FF Collection","Year 3 FF Collection","C4-Year 4 FF Collection","C5-Year 5 FF Collection" ),
                     labels=c("3-6 Months","1 Year","3 Years","4 Years", "5 Years")) +
  geom_point(aes(color = COLLECTION), alpha = 0.5, size = 3) +
  theme_classic() +
  labs(tag = "(C)")+
  theme(legend.position=c(0.2,0.9),
        text=element_text(size=14),
        legend.background = element_rect(fill="transparent"), 
        plot.title = element_text(vjust = 7, size=20),
        plot.margin = margin(0.8,0,0,0, unit="cm"),
        plot.tag.position =  c(0.01,1.075),
        plot.tag = element_text(face="bold"),)+
  xlab("PCo1")+
  ylab("PCo2")
plotall

# Plot PCoA with age color-coded CONTROL Subjects
plot_health<- plot_ordination(
  physeq = health,
  ordination = data_health,
  color = "COLLECTION",
  title = "Normal pregnancy (n=1397)"
)   + 
  scale_color_manual(values = c("#FDE725FF","#95D840FF","#1F968BFF","darkorchid4", "#FF5F15" ),
                     name="Age at Stool Sample\nCollection",
                     breaks=c("Fecal Flora Collection month 3","Year 1 FF Collection","Year 3 FF Collection","C4-Year 4 FF Collection","C5-Year 5 FF Collection" ),
                     labels=c("3-6 Months","1 Year","3 Years","4 Years", "5 Years")) +
  geom_point(aes(color = COLLECTION), alpha = 0.5, size = 3) +
  theme_classic() +
  theme(legend.position="none", plot.title = element_text(size=15))+
  xlab("Pco1")+
  ylab("Pco2")

# Plot PCoA with age color-coded PE Subjects
plot_pe <- plot_ordination(
  physeq = pe.df,
  ordination = data_pe,
  color = "COLLECTION",
  title = "Preeclampsia (n=162)"
)   + 
  scale_color_manual(values = c("#FDE725FF","#95D840FF","#1F968BFF","darkorchid4", "#FF5F15" ),
                     name="Age at Stool Sample\nCollection",
                     breaks=c("Fecal Flora Collection month 3","Year 1 FF Collection","Year 3 FF Collection","C4-Year 4 FF Collection","C5-Year 5 FF Collection" ),
                     labels=c("3-6 Months","1 Year","3 Years","4 Years", "5 Years")) +
  geom_point(aes(color = COLLECTION), alpha = 0.5, size = 3) +
  theme_classic() +
  theme(legend.position="none", plot.title = element_text(size=15))+
  xlab("Pco1")+
  ylab("Pco2")

# Plot PCoA with age color-coded SPTB Subjects
plot_sptb <- plot_ordination(
  physeq = sptb.df,
  ordination = data_sptb,
  color = "COLLECTION",
  title = "Spontaneous preterm birth (n=105) "
)   + 
  scale_color_manual(values = c("#FDE725FF","#95D840FF","#1F968BFF","darkorchid4", "#FF5F15" ),
                     name="Age at Stool Sample\nCollection",
                     breaks=c("Fecal Flora Collection month 3","Year 1 FF Collection","Year 3 FF Collection","C4-Year 4 FF Collection","C5-Year 5 FF Collection" ),
                     labels=c("3-6 Months","1 Year","3 Years","4 Years", "5 Years")) +
  geom_point(aes(color = COLLECTION), alpha = 0.5, size = 3) +
  theme_classic() +
  theme(
    legend.position="none", plot.title = element_text(size=15))+
  xlab("Pco1")+
  ylab("Pco2")

# Arrange Beta Diversity plots 
plot_beta <- ggarrange(plotall, ggarrange(plot_health, plot_pe, plot_sptb, ncol=3), nrow=2)

# Arrange Alpha and Beta Diversity plots 
plot_alpha_beta <- ggarrange(ggarrange(pe, sptb, nrow=2), plotall, ggarrange(plot_health, plot_pe, plot_sptb, nrow=3), ncol=3, 
                  widths = c(0.6, 0.9, 0.4))+
  theme(plot.margin = margin(1.5,0.1,0.1,0.1, "cm")) 
plot_alpha_beta
```

# Figure 3. MAASLIN2 Heatmap of main phenotypes, breastfeeding and mode of delivery

```{r fig3}
# Load results from MAASLIN2 analysis of main phenotypes 
mydf <- read.table("/significant_results.tsv"
                   , sep = '\t', header = TRUE)

# Subset data, removing subjects with less than 10 counts for a microbe, removing extra predictors and select significant q values
mydf <- subset(mydf, N.not.0>=10)
mydf <- subset(mydf, metadata!="age")
mydf <- subset(mydf, metadata!="cgender")
mydf <- subset(mydf, metadata!="craceeth")
mydf <- subset(mydf, metadata!="antibiotic")
mydf <- subset(mydf, qval<0.05)

# Maaslin2 heatmap
maaslin2_heatmap <-
  function(
    output_results,
    title = NA,
    cell_value = 'qval',
    data_label = 'data',
    metadata_label = 'metadata',
    border_color = 'grey93',
    color = colorRampPalette(c("blue", "grey90", "red")),
    col_rotate = 90,
    first_n = 35) {
    
    # read MaAsLin output
    df <- mydf
    
    title_additional <- ""
    
    title_additional <- ""
    if (!is.na(first_n) & first_n > 0 & first_n < dim(df)[1]) {
      if (cell_value == 'coef') {
        df <- df[order(-abs(df[[cell_value]])) , ]
      } else{
        df <- df[order(df[[cell_value]]), ]
      }
      # get the top n features with significant associations
      df_sub <- df[1:first_n,]
      for (first_n_index in seq(first_n, dim(df)[1]))
      {
        if (length(unique(df_sub$feature)) == first_n)
        {
          break
        }
        df_sub <- df[1:first_n_index,]
      }
      # get all rows that have the top N features
      df <- df[which(df$feature %in% df_sub$feature),]
      title_additional <- paste("Top", first_n, sep=" ")
    }
    
    if (dim(df)[1] < 2) {
      print('There are no associations to plot!')
      return(NULL)
    }
    
    metadata <- df$metadata
    data <- df$feature
    dfvalue <- df$value
    value <- NA
    
    # values to use for coloring the heatmap
    # and set the colorbar boundaries
    if (cell_value == "pval") {
      value <- -log(df$pval) * sign(df$coef)
      value <- pmax(-20, pmin(20, value))
      if (is.null(title))
        title <- "(-log(pval)*sign(coeff))"
    } else if (cell_value == "qval") {
      value <- -log(df$qval) * sign(df$coef)
      value <- pmax(-20, pmin(20, value))
      if (is.null(title))
        title <- "(-log(qval)*sign(coeff))"
    } else if (cell_value == "coef") {
      value <- df$coef
      if (is.null(title))
        title <- "(coeff)"
    }
    
    if (title_additional!="") {
      title <- paste(title_additional, "features with significant associations", title, sep=" ")
    } else {
      title <- paste("Significant associations", title, sep=" ")
    }
    
    # identify variables with more than one level present
    verbose_metadata <- c()
    metadata_multi_level <- c()
    for (i in unique(metadata)) {
      levels <- unique(df$value[df$metadata == i])
      if (length(levels) > 1) {
        metadata_multi_level <- c(metadata_multi_level, i)
        for (j in levels) {
          verbose_metadata <- c(verbose_metadata, paste(i, j))
        }
      } else {
        verbose_metadata <- c(verbose_metadata, i)
      }
    }
    
    n <- length(unique(data))
    m <- length(unique(verbose_metadata))
    
    if (n < 2) {
      print(
        paste(
          "There is not enough features in the associations",
          "to create a heatmap plot.",
          "Please review the associations in text output file.")
      )
      return(NULL)
    }
    
    if (m < 2) {
      print(
        paste(
          "There is not enough metadata in the associations",
          "to create a heatmap plot.",
          "Please review the associations in text output file.")
      )
      return(NULL)
    }
    
    a = matrix(0, nrow = n, ncol = m)
    a <- as.data.frame(a)
    
    rownames(a) <- unique(data)
    colnames(a) <- unique(verbose_metadata)
    
    for (i in seq_len(dim(df)[1])) {
      current_metadata <- metadata[i]
      if (current_metadata %in% metadata_multi_level) {
        current_metadata <- paste(metadata[i], dfvalue[i])
      }
      if (abs(a[as.character(data[i]), 
                as.character(current_metadata)]) > abs(value[i]))
        next
      a[as.character(data[i]), as.character(current_metadata)] <- value[i]
    }
    
    # get the range for the colorbar
    max_value <- ceiling(max(a))
    min_value <- ceiling(min(a))
    range_value <- max(c(abs(max_value),abs(min_value)))
    breaks <- seq(-1*range_value, range_value, by = 1)
    
    p <- NULL
    tryCatch({
      p <-
        pheatmap::pheatmap(
          a,
          cellwidth = 5,
          cellheight = 5,
          # changed to 3
          main = title,
          fontsize = 6,
          kmeans_k = NA,
          border = TRUE,
          show_rownames = TRUE,
          show_colnames = TRUE,
          scale = "none",
          cluster_rows = FALSE,
          cluster_cols = TRUE,
          clustering_distance_rows = "euclidean",
          clustering_distance_cols = "euclidean",
          legend = TRUE,
          border_color = border_color,
          color = color(range_value*2),
          breaks = breaks,
          treeheight_row = 0,
          treeheight_col = 0,
          display_numbers = matrix(ifelse(
            a > 0.0, "+", ifelse(a < 0.0, "-", "")), nrow(a)),
          silent = TRUE
        )
    }, error = function(err) {
      logging::logerror("Unable to plot heatmap")
      logging::logerror(err)
    })
    return(p)
  }

maaslin2_heatmap()

```

# Supplemental Table S1. Comparisons of VDAART pregnant women’s characteristics and those of their infants included and not included in this study.

```{r tabS1}

# variables 
vars=c("mage","sitename","craceeth", "antibiotic", "antimis", "income","married", "cgender", "masthma","preeclampsia","pe_and_preterm", "pe_no_preterm", "trmt", "sPTB","sptb_and_pprom","sptb_no_pprom","delgestagewks","GA_U37wks" , "GA_U34wks","Csection","m6breastfed","bfmis","pretermdelivery", "pihghtn","pprom", "grad", "bweight", "bwcat")

# plot table
tableOne <- CreateTableOne(vars = vars, strata = "microbiome", data = all806)
tableOneExport=print(tableOne, quote = F, noSpaces = TRUE, showAllLevels = T, na.rm=T)

```

# Supplemental Table S2. Associations of alpha diversity (Shannon Index) with preeclampsia (PE) and spontaneous preterm birth (sPTB). 

```{r tabS2}

### Linear regression alpha diversity associations
## Preeclampsia 
# Crude model
m3 <- glm(alpha_shan ~ preeclampsia  , data=M3)
y1 <- glm(alpha_shan ~ preeclampsia  , data=Y1)
y3 <- glm(alpha_shan ~ preeclampsia  , data=Y3)
y4 <- glm(alpha_shan ~ preeclampsia  , data=Y4)
y5 <- glm(alpha_shan ~ preeclampsia  , data=Y5)

# Estimates, p-values and 95% confidence intervals
cbind(rbind(coef(summary(m3))["preeclampsiaYes",c("Estimate","Pr(>|t|)")], coef(summary(y1))["preeclampsiaYes",c("Estimate","Pr(>|t|)")],     coef(summary(y3))["preeclampsiaYes",c("Estimate","Pr(>|t|)")], coef(summary(y4))["preeclampsiaYes",c("Estimate","Pr(>|t|)")], coef(summary(y5))["preeclampsiaYes",c("Estimate","Pr(>|t|)")]),
rbind(confint(m3)[c(2,4)], confint(y1)[c(2,4)], confint(y3)[c(2,4)], confint(y4)[c(2,4)], confint(y5)[c(2,4)]))

## SPTB
# Crude model
m3 <- glm(alpha_shan ~ sPTB  , data=M3)
y1 <- glm(alpha_shan ~ sPTB  , data=Y1)
y3 <- glm(alpha_shan ~ sPTB  , data=Y3)
y4 <- glm(alpha_shan ~ sPTB  , data=Y4)
y5 <- glm(alpha_shan ~ sPTB  , data=Y5)

# Estimates, p-values and 95% confidence intervals
cbind(rbind(coef(summary(m3))["sPTBYes",c("Estimate","Pr(>|t|)")], coef(summary(y1))["sPTBYes",c("Estimate","Pr(>|t|)")],     coef(summary(y3))["sPTBYes",c("Estimate","Pr(>|t|)")], coef(summary(y4))["sPTBYes",c("Estimate","Pr(>|t|)")], coef(summary(y5))["sPTBYes",c("Estimate","Pr(>|t|)")]),
rbind(confint(m3)[c(2,4)], confint(y1)[c(2,4)], confint(y3)[c(2,4)], confint(y4)[c(2,4)], confint(y5)[c(2,4)]))

# Adjusted models for PE and SPTB
am3 <- glm(alpha_shan ~ preeclampsia + craceeth + m6breastfed + antibiotic + cgender + sitename + age + Csection + sPTB, data=M3) # adjusted for exact age of stool sample collection as well
ay1 <- glm(alpha_shan ~ preeclampsia + craceeth + m6breastfed + antibiotic + cgender + sitename  + Csection+ sPTB , data=Y1)
ay3 <- glm(alpha_shan ~ preeclampsia + craceeth + m6breastfed + antibiotic + cgender + sitename  + Csection+ sPTB , data=Y3)
ay4 <- glm(alpha_shan ~ preeclampsia + craceeth + m6breastfed + antibiotic + cgender + sitename  + Csection+ sPTB , data=Y4)
ay5 <- glm(alpha_shan ~ preeclampsia + craceeth + m6breastfed + antibiotic + cgender + sitename  + Csection + sPTB, data=Y5)

# Estimates, p-values and 95% confidence intervals for Preeclampsia
cbind(rbind(coef(summary(am3))["preeclampsiaYes",c("Estimate","Pr(>|t|)")], coef(summary(ay1))["preeclampsiaYes",c("Estimate","Pr(>|t|)")],     coef(summary(ay3))["preeclampsiaYes",c("Estimate","Pr(>|t|)")], coef(summary(ay4))["preeclampsiaYes",c("Estimate","Pr(>|t|)")], coef(summary(ay5))["preeclampsiaYes",c("Estimate","Pr(>|t|)")]),
rbind(confint(am3)["preeclampsiaYes",c(1,2)], confint(ay1)["preeclampsiaYes",c(1,2)], confint(ay3)["preeclampsiaYes",c(1,2)], confint(ay4)["preeclampsiaYes",c(1,2)], confint(ay5)["preeclampsiaYes",c(1,2)]))

# Estimates, p-values and 95% confidence intervals for SPTB
cbind(rbind(coef(summary(am3))["sPTBYes",c("Estimate","Pr(>|t|)")], coef(summary(ay1))["sPTBYes",c("Estimate","Pr(>|t|)")],     coef(summary(ay3))["sPTBYes",c("Estimate","Pr(>|t|)")], coef(summary(ay4))["sPTBYes",c("Estimate","Pr(>|t|)")], coef(summary(ay5))["sPTBYes",c("Estimate","Pr(>|t|)")]),
rbind(confint(am3)["sPTBYes",c(1,2)], confint(ay1)["sPTBYes",c(1,2)], confint(ay3)["sPTBYes",c(1,2)], confint(ay4)["sPTBYes",c(1,2)], confint(ay5)["sPTBYes",c(1,2)]))

```

# Supplemental Table S3. Associations of beta diversity (Bray-Curtis dissimilarity) with preeclampsia (PE) and spontaneous preterm birth (sPTB). 

```{r tabS3}
## Use Adonis PERMANOVA to test for associations of beta diversity with PE and sPTB

set.seed(11)

# Remove subjects with missing values on covariates, create total (_beta) and complete case dataset (beta_cc) per time point
# total data for univariate analysis and complete case for adjusted analysis 
# 3-6 months
age0.5_beta <- age0.5
age0.5_beta_cc <- subset_samples(age0.5_beta, m6breastfed!="")
age0.5_beta_cc <- subset_samples(age0.5_beta_cc, antibiotic!="")
# year 1
age1_beta <- age1
age1_beta_cc <- subset_samples(age1_beta, antibiotic!="")
age1_beta_cc <- subset_samples(age1_beta_cc, m6breastfed!="")
# year 3
age3_beta <- age3
age3_beta_cc <- subset_samples(age3_beta, antibiotic!="")
age3_beta_cc <- subset_samples(age3_beta_cc, m6breastfed!="")
# year 4
age4_beta <- age4
age4_beta_cc <- subset_samples(age4_beta, antibiotic!="")
age4_beta_cc <- subset_samples(age4_beta_cc, m6breastfed!="")
# year 5
age5_beta <- age5
age5_beta_cc <- subset_samples(age5_beta, antibiotic!="")
age5_beta_cc <- subset_samples(age5_beta_cc, m6breastfed!="")

# Calculate bray curtis distance matrix for total data and complete case data (_cc)
# 3-6 months 
age0.5_bray <- phyloseq::distance(age0.5_beta, method = "bray")
age0.5_bray_cc <- phyloseq::distance(age0.5_beta_cc, method = "bray")
# Create metadata data frame 
sampledf0.5 <- data.frame(sample_data(age0.5_beta))
sampledf0.5_cc <- data.frame(sample_data(age0.5_beta_cc))

# Calculate bray curtis distance matrix
# year 1
age1_bray <- phyloseq::distance(age1_beta, method = "bray")
age1_bray_cc <- phyloseq::distance(age1_beta_cc, method = "bray")
# Create metadata data frame
sampledf1 <- data.frame(sample_data(age1_beta))
sampledf1_cc <- data.frame(sample_data(age1_beta_cc))

# Calculate bray curtis distance matrix
# year 3
age3_bray <- phyloseq::distance(age3_beta, method = "bray")
age3_bray_cc <- phyloseq::distance(age3_beta_cc, method = "bray")
# Create metadata data frame
sampledf3 <- data.frame(sample_data(age3_beta))
sampledf3_cc <- data.frame(sample_data(age3_beta_cc))

# Calculate bray curtis distance matrix
# year 4
age4_bray <- phyloseq::distance(age4_beta, method = "bray")
age4_bray_cc <- phyloseq::distance(age4_beta_cc, method = "bray")
# Create metadata data frame
sampledf4 <- data.frame(sample_data(age4_beta))
sampledf4_cc <- data.frame(sample_data(age4_beta_cc))

# Calculate bray curtis distance matrix
# year 5
age5_bray <- phyloseq::distance(age5_beta, method = "bray")
age5_bray_cc <- phyloseq::distance(age5_beta_cc, method = "bray")
# Create metadata data frame
sampledf5 <- data.frame(sample_data(age5_beta))
sampledf5_cc <- data.frame(sample_data(age5_beta_cc))

# Adonis test for associations between bray beta diversity and PE univariate (total data)
adonis2(age0.5_bray ~ preeclampsia  , data = sampledf0.5) 
adonis2(age1_bray ~ preeclampsia , data = sampledf1)  
adonis2(age3_bray ~ preeclampsia , data = sampledf3) 
adonis2(age4_bray ~ preeclampsia , data = sampledf4) 
adonis2(age5_bray ~ preeclampsia , data = sampledf5) 

# Adonis test for associations between bray beta diveristy and PE with covariates_cc
adonis2(age0.5_bray_cc ~ preeclampsia + age + m6breastfed + cgender + 
          Csection + antibiotic + craceeth +sPTB + sitename , data = sampledf0.5_cc)    
adonis2(age1_bray_cc ~ preeclampsia+ m6breastfed + cgender + 
          Csection + antibiotic + craceeth+sPTB + sitename, data = sampledf1_cc) 
adonis2(age3_bray_cc ~ preeclampsia+ m6breastfed + cgender + 
          Csection + antibiotic + craceeth+sPTB + sitename, data = sampledf3_cc) 
adonis2(age4_bray_cc ~ preeclampsia+ m6breastfed + cgender + 
          Csection + antibiotic + craceeth+sPTB + sitename, data = sampledf4_cc) 
adonis2(age5_bray_cc ~ preeclampsia+ m6breastfed + cgender + 
          Csection + antibiotic + craceeth+sPTB + sitename, data = sampledf5_cc) 

# Adonis test for associations between bray beta diversity and SPTB univariate (total data)
adonis2(age0.5_bray ~ sPTB  , data = sampledf0.5) 
adonis2(age1_bray ~ sPTB , data = sampledf1)  
adonis2(age3_bray ~ sPTB , data = sampledf3) 
adonis2(age4_bray ~ sPTB , data = sampledf4) 
adonis2(age5_bray ~ sPTB , data = sampledf5) 

# Adonis test for associations between bray beta diveristy and SPTB with covariates_cc
adonis2(age0.5_bray_cc ~ sPTB + age + m6breastfed + cgender + 
          Csection + antibiotic + craceeth +preeclampsia + sitename , data = sampledf0.5_cc)    
adonis2(age1_bray_cc ~ sPTB+ m6breastfed + cgender + 
          Csection + antibiotic + craceeth+preeclampsia + sitename, data = sampledf1_cc) 
adonis2(age3_bray_cc ~ sPTB+ m6breastfed + cgender + 
          Csection + antibiotic + craceeth+preeclampsia + sitename, data = sampledf3_cc) 
adonis2(age4_bray_cc ~ sPTB+ m6breastfed + cgender + 
          Csection + antibiotic + craceeth+preeclampsia + sitename, data = sampledf4_cc) 
adonis2(age5_bray_cc ~ sPTB+ m6breastfed + cgender + 
          Csection + antibiotic + craceeth+preeclampsia + sitename, data = sampledf5_cc) 

```


# Supplemental Figure S1. Data intersections across all timepoints.  

```{r figS1}

all_unique$'3-6 Months (n=265)' <- ifelse(all_unique$vid %in% db36$vid, 1, 0)
all_unique$'1 Year (n=436)' <- ifelse(all_unique$vid %in% db1$vid, 1, 0)
all_unique$'3 Years (n=506)' <- ifelse(all_unique$vid %in% db3$vid, 1, 0)
all_unique$'4 Years (n=313)' <- ifelse(all_unique$vid %in% db4$vid, 1, 0)
all_unique$'5 Years (n=151)' <- ifelse(all_unique$vid %in% db5$vid, 1, 0)

int <- colnames(all_unique)[442:438]

Upsetplot <- upset(all_unique, intersect = int, name="Intersection Pattern",width_ratio=0.1,min_size=10,base_annotations=list(),
      sort_sets=F, sort_intersections_by = "degree", encode_sets=FALSE, set_sizes = F,
      matrix=(intersection_matrix() + 
                annotate(geom='text', label='45', x=14, y=5.4, 
                                               size=5 ) +
  annotate(geom='text', label='64', x=13, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='20', x=12, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='38', x=11, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='50', x=10, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='20', x=9, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='24', x=8, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='57', x=7, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='59', x=6, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='16', x=5, y=5.4, 
           size=5 ) +
  annotate(geom='text', label='50', x=4, y=5.4, 
           size=5)+
    annotate(geom='text', label='30', x=3, y=5.4, 
             size=5)+
    annotate(geom='text', label='106', x=2, y=5.4, 
             size=5)+
    annotate(geom='text', label='34', x=1, y=5.4, 
             size=5)+
    annotate(geom='text', label='n=', x=0.5, y=5.4, 
             size=5) ))+
  geom_linerange(xmin = 0.8, xmax = 1.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 1.8, xmax = 2.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 2.8, xmax = 3.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 3.8, xmax = 4.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 4.8, xmax = 5.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 5.8, xmax = 6.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 6.8, xmax = 7.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 7.8, xmax = 8.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 8.8, xmax = 9.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 9.8, xmax = 10.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 10.8, xmax = 11.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 11.8, xmax = 12.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 12.8, xmax = 13.2 , y=5.3,  color = 'grey90')+
  geom_linerange(xmin = 13.8, xmax = 14.2 , y=5.3,  color = 'grey90')+
  ggtitle('Longitudinal Fecal Microbiome sampling intersections (n=672)')+
  theme(plot.margin = unit(c(1,1,1,1), "cm"))+
  theme(plot.title = element_text(vjust = 10))+
  coord_cartesian(ylim = c(1, 4.7), clip = "off")+
  expand_limits(x=14.8)
Upsetplot
```

# Supplemental Figure S2. Infant gut microbiome diversity trajectories from birth to age 5 years color coded for breastfeeding and mode of delivery.

```{r figS2}
# Select only subjects with breastfeeding (m6breastfed) data   
breastcc <- dfall[!is.na(dfall$m6breastfed),]
breastcc$m6breastfed <- factor(breastcc$m6breastfed, levels = c("Yes", "No"))

# Create the breastfeeding plot for PE
plot1 <- ggplot(breastcc[breastcc$alpha_shan!=0,], aes(x = Time, y = alpha_shan, col = m6breastfed)) +
  geom_boxplot() +
  facet_wrap(~preeclampsia, labeller = as_labeller(c(`No` = "Preeclampsia: No", `Yes` = "Preeclampsia: Yes"))) +
  scale_color_discrete(name = "Early Life Breastfeeding", labels = c("Yes", "No")) +
  xlab("Offspring Age") +
  ylab("Shannon Index") +
  theme_bw() +
  theme(axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17),
        axis.text.x = element_text(size = 12),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 20),
        strip.text = element_text(size = 17))

# Create the breastfeeding plot for SPTB
plot2 <- ggplot(breastcc[breastcc$alpha_shan!=0,], aes(x = Time, y = alpha_shan, col = m6breastfed)) +
  geom_boxplot() +
  facet_wrap(~sPTB, labeller = as_labeller(c(`No` = "Spontaneous Preterm Birth: No", `Yes` = "Spontaneous Preterm Birth: Yes"))) +
  scale_color_discrete(name = "Early Life Breastfeeding", labels = c("Yes", "No")) +
  xlab("Offspring Age") +
  ylab("Shannon Index") +
  theme_bw() +
  theme(axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17),
        axis.text.x = element_text(size = 12),
        legend.position = "none",
        strip.text = element_text(size = 17))

# Select only subjects with mode of delivery (Csection) data   
cseccc <- dfall[!is.na(dfall$Csection),]
cseccc$Csection <- factor(cseccc$Csection, levels = c("No", "Yes"))

# Create the Csection plot for PE
plot3 <- ggplot(cseccc[cseccc$alpha_shan!=0,], aes(x = Time, y = alpha_shan, col = Csection)) +
  geom_boxplot() +
  facet_wrap(~preeclampsia, labeller = as_labeller(c(`No` = "Preeclampsia: No", `Yes` = "Preeclampsia: Yes"))) +
  xlab("Offspring Age") +
  ylab("Shannon Index") +
  theme_bw() +
  theme(axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17),
        axis.text.x = element_text(size = 12),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 20),
        strip.text = element_text(size = 17))

# Create the Csection plot for PE
plot4 <- ggplot(cseccc[cseccc$alpha_shan!=0,], aes(x = Time, y = alpha_shan, col = Csection)) +
  geom_boxplot() +
  facet_wrap(~sPTB, labeller = as_labeller(c(`No` = "Spontaneous Preterm Birth: No", `Yes` = "Spontaneous Preterm Birth: Yes"))) +
  xlab("Offspring Age") +
  ylab("Shannon Index") +
  theme_bw() +
  theme(axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17),
        axis.text.x = element_text(size = 12),
        legend.position = "none",
        strip.text = element_text(size = 17))

plot3 <- plot3 + scale_color_brewer(palette = "Dark2", name = "Vaginal Delivery", labels = c("Yes", "No"))
plot4 <- plot4 + scale_color_brewer(palette = "Dark2", name = "Vaginal Delivery", labels = c("Yes", "No"))

#arrange grid
grid1 <- ggarrange(plot1, plot2, ncol=2, common.legend = TRUE, legend = "top", labels = c("A", "B"),
                   font.label = list(size=17))
grid2 <- ggarrange(plot3, plot4, ncol=2, common.legend = TRUE, legend = "top", labels = c("C", "D"),
                   font.label = list(size=17))

plot_final <- grid.arrange(grid1, grid2, nrow=2)
plot_final
```
